Stability and Performance of Coalitions of Prosumers Through Diversification in the Smart Grid

### 2015-11-04 Appendix.tex
figures sizes
update

\begin{figure}[b]
\includegraphics[scale=.5]{search}
\caption{Evolution of the global utility $ \mathcal{U}(CS) $ (blue diamond curve, left axis) and the number of agents involved in the coalitions (red circle curve, right axis) during the greedy optimization of algorithm \ref{alg:algo1} }
\label{fig:search}
\end{figure}

@@

\subsection{Net production series}\label{appendixB}

@@

>>>>

\begin{figure}[b]
\includegraphics[scale=.5]{./figs/figure_7}
\caption{Evolution of the global utility $ \mathcal{U}(CS) $ (blue diamond curve, left axis) and the number of agents involved in the coalitions (red circle curve, right axis) during the greedy optimization of algorithm \ref{alg:algo1} }
\label{fig:search}
\end{figure}

@@

\subsection{Net production series}
\label{appendixB}

@@

\subsection{Resilience algorithm}
\label{appendix_resilience}
The following algorithm (see algorithm \ref{alg:algo4}) explains the process through which the resilience of a coalition structure is explored. For a given coalition structure, we consider random failures of physical components of the agents. This results in the agent being unable to support its coalition. Therefore, the agent is simply removed from the coalition, which impacts the coalition's production. Eventually, after some number of agent failures, some coalitions might not be sufficiently stable and productive to fulfill the market's constraints. In this case, the coalition fails, and is removed from the market.
\begin{algorithm}
	\KwData{$ CS = \left\{ S_1,S_2,...,S_k \right\} $ Coalition structure \;}

	$ pool = \cup_{k \in N_{COAL}} S_k $ \; 
	\While{ $ pool \neq \emptyset $}{
		Select randomly agent i from pool\;
		i fails $ \Rightarrow $ Remove i from its coalition : $ S_k \leftarrow S_k - \{i\} $ \;
		$ pool = pool - \{i\} $\;
		\If{ $ Pr \left[ P_{S_k} < P^{MIN} \right] > \phi $}{
			$S_k$ fails $ \Rightarrow $ Remove $S_k$ from the market
			} 
		}
	\caption{Random failures algorithm}
\label{alg:algo4}
\end{algorithm}

### 2015-11-04 Article.tex
update

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Authors
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Nicolas Gensollen, Vincent Gauthier, Monique Becker and Michel Marot}

@@

More precisely, in our model, any entity that wishes to sell its production on some energy market has to estimate and announce a production capacity for an upcoming period of time. If a contract is bought, the entity commits to injecting exactly, at any time of the contracted period, the announced amount of power, under financial penalties if it fails. Since prosumers use exclusively renewables, contracts come naturally with some amount of risk due to the intermittent nature of most renewables. Using storage as buffers in order to reduce risks is a popular approach. Actually, a whole branch of the smart grid literature is even studying the possibility to use electric vehicles as moving storage capacities for stabilizing the grid. Without storage and proper control, the over-producing state of the agents, and thus the power they inject in the grid, might be rather unstable. This is clearly unacceptable for the grid operator that cannot ensure system stability if it has to deal with numerous small unpredictable entities.

Forming coalitions/aggregations is an envisaged solution to both the number of entities soliciting the operator, and their high variability. Indeed, it has been argued that using a multi-level aggregators architecture for the control could lead to better performance on the communication side \cite{Negeri2012}, since an aggregator can be considered as a single node by the level above it. On the other hand, it is well-known that diversification of the assets is a way of minimizing risks when constructing a portfolio. One thus expect a more stable and predictable energy production for a coalition than for a single agent. Nevertheless, all coalitions are obviously not equally stable or productive, which means that special attention should be paid to the aggregation step. Recall that prosumers have both a consumption and a production component, each depending on location and time, meaning that there are complex underlying correlations between the agents. This is a central topic of the present paper : given N prosumers, what coalitions should be formed so that the compromise between expected production and variability is optimized ?

We will see that the variability in the aggregated productions can be quantified to a certain extent by the correlation among the agents forming the coalitions. Understanding the correlation relationships among the agents can thus give an indication about what coalitions to form and how much they should sell. More precisely, we use a graph representation of the correlation structure to gain insight about the expected production to risk ratios of different coalitions. We build a framework in which the system operator specifies both the minimum production acceptable to enter the market and restrain the amount of risk he is willing to take. We then propose a graph heuristic that uses decorrelated cliques in order to form diversified productive and stable coalitions, and we compare the results with other formation strategies (see section \ref{sec:results}). Rather than maximizing the profit of each agent in the grid (prosumers and grid operator), our algorithm tries to form the most productive coalitions given a maximum amount of risk acceptable. In other words, it tries to maximize coalitions profits without considering individual retributions to single agents (a task often studied through game theory). Furthermore, the pricing strategies for both the prosumer and the grid is beyond the scope of this paper.

@@

The paper is organized as follows, section \ref{sec:related} gives a brief overview of the related literature, section \ref{sec:data} clarifies how we generated realistic prosumer production traces based on weather data. In section \ref{sec:notations}, we define most of the notations and explain why correlation between prosumers is a quantity of interest for our objective. Based on the conclusions of section \ref{sec:notations}, we develop in section \ref{sec:utility} a utility function quantifying how much power a coalition can announce on the market given an accepted risk level. We then develop, in section \ref{sec:forming} a greedy optimization algorithm that uses decorrelated cliques as inputs and improves their utility over a correlation-constrained environment. Finally, section \ref{sec:results} provides some results both on performance of the method and resilience of the coalitions formed.

@@

As explained in the section above, allowing entities using mainly renewables to inject power in the grid is a difficult challenge. Indeed, on the contrary to fossil plants whose production can be scheduled in advance to meet the expected consumption, renewables by definition only produce when the resource is present. Unfortunately, these moments do not necessarily coincide with the consumption peak hours \cite{Milligan2010}. One possible solution consists in shifting some of the loads towards high productivity periods. Demand side management techniques can indeed mitigate the consumption peaks and wast less production \cite{Milligan2010} \cite{Logenthiran2012}. Since end users tend to seek a maximum utility at a minimum cost, dynamic pricing appears as a good way to give incentives to them. By carefully scheduling the prices, it is assumed that the load curve can be shaped to some extend.

@@

An essential component of the smart grid is the smart meter which makes the interface between the end user and the rest of the system. Smart meters coupled with sensors measure quantities of interest (like instantaneous consumption), receive informations from the grid (electricity prices for instance), and take actions accordingly (demand side management programs). Smart meters are currently and gradually deployed, and will probably provide interesting datasets to work on. Unfortunately, at the time this paper was written, production and consumption data for prosumers over a large region were not yet available to our knowledge. Some interesting experiments are nonetheless being conducted and data are progressively made public \cite{ISSDA}. In this paper, we use weather quantities like wind speed or solar radiance as alternative data for generating realistic production and consumption series. Fortunately, these kinds of data are easier to find, and since the development of small personal weather stations, their geographical granularity keeps increasing.  Since these quantities depend both on time and location, we discretize time into slots and space into zones in the following. A zone is simply a portion of the considered region of study for which we sampled data. Therefore, if prosumers i and j are positioned on the same zone, they are exposed to the same weather. Adding some intra-zone noise can easily be done though not considered in this paper. More formally, we denote by $ P_{i}(t) $ the instantaneous extra-production of agent i at time t :

@@

\begin{figure}
\label{fig:process}
	\includegraphics[scale=.5]{Fig2}
	\caption{Process.}
\end{figure}

@@

Suppose now that coalition S has to suggest a production value $ P_{S}^{CRCT} $ to enter the market. This means that, during the time S is on the market, it will have to inject in the grid exactly $ P_{S}^{CRCT} $ at any time t and will be rewarded proportionally to this amount, with penalties if it deviates. Obviously, the actual extra-production will not be constant at this value and will oscillate due to intermittences in the production and consumption. If S always produces more than $ P_{S}^{CRCT} $, it will never have to pay penalties, but it is losing some gains since it could have announced a higher contract value. If the production oscillates around $ P_{S}^{CRCT} $, by using batteries or demand side management techniques, S could be able to maintain its production to the contract value at any time. Nevertheless, if the oscillations are too important compared to the available storage capacity, S will probably break the contract and pay penalties. We can see that there is a return over risk trade-off here. Coalitions should thus find the right balance between announcing too low and losing some potential gains, and claiming too high and paying penalties.

@@

We consider that prosumers fail randomly, and we denote by $ \psi \in [0,1] $ the fraction of agents that failed. Figure \ref{fig:resilience} exhibits how the resilience of the coalition structures evolves according to $ \psi $. On the top subplot, $ P^{MIN} $ was voluntarily selected relatively low such that the resiliences of the three structures fit on the same figure. When the $ P^{MIN} $ requirement increases, the differences between the algorithms also increase as visible on the bottom subplot of figure \ref{fig:resilience}. The decorrelated coalitions achieve a more resilient production on the market in the sense that they sustain a higher fraction of node failures.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Authors
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author{Nicolas Gensollen, Vincent Gauthier, Monique Becker and Michel Marot \\
\IEEEauthorblockA{CNRS UMR 5157 SAMOVAR, \\
Telecom SudParis/Institut Mines Telecom\\
Email: \{nicolas.gensollen, vincent.gauthier, michel.marot, monique.becker\}@telecom-sudparis.eu}}

@@

More precisely, in our model, any entity that wishes to sell its production on some energy market has to estimate and announce a production capacity for an upcoming period of time. If a contract is bought, the entity commits to injecting, at any time of the contracted period, the announced amount of power, under financial penalties if it fails. Since prosumers use exclusively renewables, contracts come naturally with some amount of risk due to the intermittent nature of most renewables. Using storage as buffers as to reduce risks is a popular approach. Without storage and proper control, the over-producing state of the agents, and thus the power they inject in the grid, might be rather unstable. This is unacceptable for the grid operator that cannot ensure system stability if it has to deal with numerous small unpredictable entities.

Forming coalitions/aggregations is an envisaged solution to both the number of entities soliciting the operator, and their high variability. Indeed, it has been argued that using a multi-level aggregators architecture for the control could lead to better performance on the communication side \cite{Negeri2012}. On the other hand, it is known that diversification of the assets is a way of minimizing risks when building a portfolio. One thus expect a more stable and predictable energy production for a coalition than for a single agent. Nevertheless, all coalitions are not equally stable or productive. Special attention should thus be paid to the aggregation step. Recall that prosumers have both a consumption and a production component, each depending on location and time, meaning that there are complex underlying correlations between the agents. This is a central topic of the present paper : given N prosumers, what coalitions should be formed so that the compromise between expected production and variability is optimized ?

We will see that the variability in the aggregated productions can be quantified to a certain extent by the correlation among the agents forming the coalitions. Understanding the correlation relationships among the agents can thus give an indication about what coalitions to form and how much they should sell. More precisely, we use a graph representation of the correlation structure to gain insight about the expected production to risk ratios of different coalitions. We build a framework in which the system operator specifies both the minimum production acceptable to enter the market and restrain the amount of risk he is willing to take. We then propose a graph heuristic that uses decorrelated cliques in order to form diversified productive and stable coalitions, and we compare the results with other formation strategies (see section \ref{sec:results}). Rather than maximizing the profit of each agent in the grid (prosumers and grid operator), our algorithm tries to form the most productive coalitions given a maximum amount of risk acceptable. In other words, it tries to maximize coalitions profits without considering individual retributions to single agents. Furthermore, the pricing strategies for both the prosumer and the grid is beyond the scope of this paper.

@@

The paper is organized as follows, section \ref{sec:related} gives a brief overview of the related literature, section \ref{sec:data} clarifies how we generated realistic prosumer production traces based on weather data. In section \ref{sec:notations}, we define most of the notations and explain why correlation between prosumers is a quantity of interest for our objective. Based on the conclusions of section \ref{sec:notations}, we develop in section \ref{sec:utility} a utility function quantifying how much power a coalition can announce on the market given an accepted risk level. We then develop, in section \ref{sec:forming} a greedy optimization algorithm that uses decorrelated cliques as inputs and improves their utility over a correlation-constrained environment. Finally, section \ref{sec:results} provides some results both on performance of the method and resilience of the coalitions formed. The overall process is illustrated on figure \ref{fig:process}.

@@

Allowing entities using mainly renewables to inject power in the grid is a difficult challenge. On the contrary to fossil plants whose production can be scheduled in advance to meet the expected consumption, renewables only produce when the resource is present. Unfortunately, these moments do not necessarily coincide with the consumption peak hours \cite{Milligan2010}. One possible solution consists in shifting some of the loads towards high productivity periods. Demand side management techniques can indeed mitigate the consumption peaks and waste less production \cite{Milligan2010} \cite{Logenthiran2012}. Since end users tend to seek a maximum utility at a minimum cost, dynamic pricing appears as a good way to give incentives to them. By carefully scheduling the prices, it is assumed that the load curve can be shaped to some extend.

@@

\begin{figure}
\label{fig:process}
	\includegraphics[scale=.5]{Fig2}
	\caption{Process}
\end{figure}

An essential component of the smart grid is the smart meter which makes the interface between the end user and the rest of the system. Smart meters coupled with sensors measure quantities of interest (like instantaneous consumption), receive informations from the grid (electricity prices for instance), and take actions accordingly (demand side management programs). Smart meters are currently and gradually deployed, and will probably provide interesting datasets to work on. Unfortunately, at the time this paper was written, production and consumption data for prosumers over a large region were not yet available to our knowledge. Some interesting experiments are nonetheless being conducted and data are progressively made public \cite{ISSDA}. In this paper, we use weather quantities like wind speed or solar radiance as alternative data for generating realistic production and consumption series. Fortunately, these kinds of data are easier to find, and since the development of small personal weather stations, their geographical granularity keeps increasing.  Since these quantities depend both on time and location, we discretize time into slots and space into zones in the following (see block 1 of fig. \ref{fig:process}). A zone is simply a portion of the considered region of study for which we sampled data. Therefore, if prosumers i and j are positioned on the same zone, they are exposed to the same weather. Adding some intra-zone noise can easily be done though not considered in this paper. More formally, we denote by $ P_{i}(t) $ the instantaneous extra-production of agent i at time t :

@@

@@

Suppose now that coalition S has to suggest a production value $ P_{S}^{CRCT} $ to enter the market. This means that, during the time S is on the market, it will have to inject in the grid exactly $ P_{S}^{CRCT} $ at any time t and will be rewarded proportionally to this amount, with penalties if it deviates. Obviously, the actual extra-production will not be constant at this value and will oscillate due to intermittences in the production and consumption. If S has an available production always greater than $ P_{S}^{CRCT} $, it is losing some gains since it could have announced a higher contract value. If the production oscillates around $ P_{S}^{CRCT} $, by using batteries or demand side management techniques, S could be able to maintain its production to the contract value at any time. Nevertheless, if the oscillations are too important compared to the available storage capacity, S will probably break the contract and pay penalties. We can see that there is a return over risk trade-off here. Coalitions should thus find the right balance between announcing too low and losing some potential gains, and claiming too high and paying penalties.

@@

We consider that prosumers fail randomly, and we denote by $ \psi \in [0,1] $ the fraction of agents that failed. Figure \ref{fig:resilience} exhibits how the resilience of the coalition structures evolves according to $ \psi $ (see appendix section \textit{Resilience algorithm} for more details). On the top subplot, $ P^{MIN} $ was voluntarily selected relatively low such that the resiliences of the three structures fit on the same figure. When the $ P^{MIN} $ requirement increases, the differences between the algorithms also increase as visible on the bottom subplot of figure \ref{fig:resilience}. The decorrelated coalitions achieve a more resilient production on the market in the sense that they sustain a higher fraction of node failures.
